---
title: "NYPD Shooting Incident Data Analysis"
author: "Maciej Dwornik"
date: "2023-09-22"
output:
  github_document: default
---

```{r libraries}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(dplyr)
```
I used the NYPD Shooting Incident Data as well as another data set regarding NYC population to analyze which borough had the most incidents, as well as how many incidents there are relative to the population.
```{r data}
data = read_csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD")
pop_data = read_csv("https://data.cityofnewyork.us/api/views/xywu-7bv9/rows.csv?accessType=DOWNLOAD")
```
## Cleaning, Analyzing, and Visualizing the Data

From the NYPD Shooting Incident Data I derived the number of incidents from each borough. We can see that Brooklyn had the most total incidents, followed by Bronx, Queens, Manhattan, and Staten Island. No data cleaning was necessary in this step.

```{r Borough data}
table(data$BORO)
barplot(table(data$BORO),
  xlab = "Borough", 
  ylab = "Incidents",
  main = "Incidents by Borough",
  cex.names = 0.8)
```

I knew each borough was not the same size and did not have the same population, so I decided to find population data of each borough to compare which borough has the most incidents per 100,000 residents.

Here we can see that Bronx actually had the most incidents per 100,000 residents, even though Brooklyn had the most incidents total (as per previous graph). Queens and Manhattan also swapped spots. Queens had more incidents total but Manhattan had more incidents per 100,000 residents.

In this step there was a bit of data wrangling I had to do. Firstly, the column names were not working so I had to rename them. The data types were also problematic and had to be changed. Another issue that came up was that the borough names in the population data were in rows, while in the other data set they were in columns so I had to switch them around.

```{r pop_data}
colnames(pop_data)[2] = "BORO"
pop_data = pop_data %>%
  rename(`2020` = "2020") %>%
  mutate(`2020` = as.numeric(`2020`)) %>%
  select(BORO, `2020`)

incidents_per_boro = table(data$BORO)
pop_data_cols = pop_data %>%
  pivot_wider(names_from = BORO, values_from = `2020`)
colnames(pop_data_cols)[1] = "TOTAL"
pop_data_cols = pop_data_cols %>%
  select(-"TOTAL")

pop_values = unlist(pop_data_cols)

incidents_per_pop = data.frame(
  Borough = names(pop_data_cols),
  Incidents = as.numeric(incidents_per_boro),
  Population = pop_values
)

incidents_per_pop$IncidentsPer100k <- incidents_per_pop$Incidents / (incidents_per_pop$Population / 100000)

print(incidents_per_pop)

barplot(incidents_per_pop$IncidentsPer100k, 
  names.arg = incidents_per_pop$Borough, 
  xlab = "Borough", 
  ylab = "Incidents per 100,000",
  main = "Incidents per 100,000 by Borough")
```

I wanted to further analyze this data and see how it differs from year to year.

```{r Analyzing the data by year}
data$OCCUR_DATE = as.character(data$OCCUR_DATE)
date_split = strsplit(data$OCCUR_DATE, "/")
data$Year = as.integer(sapply(date_split, function(x) x[3]))

incident_count = data %>%
  group_by(BORO, Year) %>%
  summarize(Incidents = n())

ggplot(incident_count, aes(x = Year, y = Incidents, fill = BORO)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "NYPD Shooting Incidents by Borough and Year",
    x = "Year",
    y = "Incident Count",
    fill = "Borough"
  )
```

Here we do the same thing we did with the initial analysis, so we use the number of incidents (this time it is incidents per year) and calculate how many incidents happened per 100,000 residents of each borough. It seems that the data is pretty consistent throughout the years, with Bronx once again having the most incidents per 100,000 residents, while Staten Island and Queens have the least.

```{r Per year per 100k}
pop_data <- pop_data %>%
  filter(BORO != "NYC Total") %>%
  mutate(BORO = toupper(BORO))
data_with_pop = data %>%
  left_join(pop_data, by = c("BORO" = "BORO")) %>%
  group_by(BORO, Year) %>%
  reframe(Incidents = n(), Population = `2020`) %>%
  mutate(IncidentsPer100k = (Incidents / (Population / 100000)))

ggplot(data_with_pop, aes(x = Year, y = IncidentsPer100k, fill = BORO)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "NYPD Shooting Incidents per 100,000 Residents by Borough and Year",
    x = "Year",
    y = "Incidents per 100,000 Residents",
    fill = "Borough"
  )
```

## Bias
There is a potential bias problem here because the shooting incident data spans over many years, however, I only used the 2020 population data. The population of course changes over the years, as well as different boroughs having different amounts of incidents per year. I could have used incident data from just one year for the initial analysis, however, I think having more data to work with was more important. For the per year data I could have possibly used different population data for different years, but the population data I had was only for each decade. I could have technically used the 2010 population data for years that were closer to 2010, and 2020 data for years that were closer to 2020 but I chose not to for efficiency's sake.

## Model

I also constructed a model predicting incidents per 100,000 residents. The blue bar represents the original data, while the red bar represents the prediction model. We can see that in the Bronx and Brooklyn boroughs the actual number of incidents exceeds the predicted number of incidents, while the rest of the boroughs the number of incidents is lower than expected.

```{r mod}  
mod = lm(IncidentsPer100k ~ Population, data=incidents_per_pop)
summary(mod)
incidents_per_pop = incidents_per_pop %>% 
  mutate(Prediction = predict(mod))

incidents_per_pop %>%
  ggplot(aes(x = Borough, y = IncidentsPer100k)) +
  geom_bar(stat = "identity", fill = "blue", alpha = 0.5) +
  geom_bar(aes(x = Borough, y = Prediction), stat = "identity", fill = "red", alpha = 0.5) +
  labs(
  title = "Actual vs. Predicted Incidents per 100,000 by Borough",
  x = "Borough",
  y = "Incidents per 100,000"
)
  
incidents_per_pop %>%
  select(-Borough)
```

## Conclusion

In conclusion, the data suggests that even though Brooklyn had the most incidents, the Bronx actually had the most incidents per 100,000 residents. We also saw that Queens and Staten Island were on the lower end of incidents per 100,000 residents, even though Queens was third overall in total incidents.